Internal models of type theory

### 2013-11-11 article-models-defn.tex
removed commented parts

\subsection{Morphisms of internal models}
\begin{defn}\label{defn:premodel-morphism}
A morphism $f:\mfM\to \mfN$ of internal models consists of
\begin{enumerate}
\item a function $\ctx(f):\ctx(\mfM)\to\ctx(\mfN)$. The function $\ctx(f)$ is also
denoted by $f_0$.
\item a morphism $\mftyp{f}{\Gamma}:\mftyp{\mfM}{\Gamma}\to\mftyp{\mfN}{f_0(\Gamma)}$ of internal models for every
$\Gamma:\ctx(M)$.
\item the judgmental equality
\begin{equation*}
\mftyp{\mftyp{f}{\Gamma}}{A}\jdeq\mftyp{f}{\ctxext{\Gamma}{A}}
\end{equation*}
\item a function $\terms[f]{A}:\terms[\mfM]{A}\to\terms[\mfN]{\mftyp{f}{\Gamma}_0(A)}$ for
every type $A$ in context $\Gamma$. 
\item preservation of context extension: 
\begin{align*}
\alpha^f_0 & :\id{f\circ\tfext^\Gamma}{\tfext^{f_0(\Gamma)}\circ\mftyp{f}{\Gamma}}\\
\alpha^f_1 & :\id{\mftyp{f}{\Gamma}_0\circ\tfext^A}{\tfext^{\mftyp{f}{\Gamma}_0(A)}\circ\mftyp{f}{\ctxext{\Gamma}{A}}}.
\end{align*}
\begin{comment}
the judgmental equality
\begin{equation*}
f_0(\ctxext{\Gamma}{A})\jdeq\ctxext{f_0(\Gamma)}\mftyp{f}{\Gamma}_0(A)
\end{equation*}
for every type $A$ in context $\Gamma$.
\end{comment}
\item preservation of weakening: 
\begin{align*}
\beta^f_0 & :\id{\mftyp{f}{\Gamma}\circ\tfwk^\Gamma}{\tfwk^{f_0(\Gamma)}\circ f}\\
\beta^f_1 & :\id{\mftyp{f}{\ctxext{\Gamma}{A}}\circ\tfwk^A}{\tfwk^\protect{\mftyp{f}{\Gamma}_0(A)}\circ\mftyp{f}{\Gamma}}.
\end{align*}
\begin{comment}
This gives the following three judgmental equalities:
\begin{enumerate}
\item the judgmental equality
\begin{equation*}
\mftyp{f}{\ctxext{\Gamma}{A}}_0(\ctxwk{A}{B})\jdeq\ctxwk{\mftyp{f}{\Gamma}_0(A)}{\mftyp{f}{\Gamma}_0(B)}
\end{equation*}
for every two types $A$ and $B$ in context $\Gamma$.
\item the judgmental equality
\begin{equation*}
\mftyp{f}{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}\circ\tfwk^A\jdeq\tfwk^{\protect{\mftyp{f}{\Gamma}_0(A)}}\circ\mftyp{f}{\ctxext{\Gamma}{B}}
\end{equation*}
\item the judgmental equality
\begin{equation*}
\terms[f]{\ctxwk{A}{B}}{{A}{y}}\jdeq\ctxwk{\mftyp{f}\Gamma_0(A)}{(\terms[f]{B}(y))}
\end{equation*}
\end{enumerate}
\end{comment}
\item preservation of substitution: 
\begin{equation*}
\gamma^f:\id{\mftyp{f}\Gamma_0\circ\tfsubst^x}{\tfsubst^\protect{\terms[f]{A}(x)}\circ\mftyp{f}{\ctxext{\Gamma}{A}}}.
\end{equation*}
\begin{comment}
\begin{enumerate}
\item the judgmental equality
\begin{equation*}
\mftyp{f}\Gamma_0(\subst{x}{P})\jdeq\subst{\terms[f]{A}(x)}{\mftyp{f}{\ctxext{\Gamma}{A}}_0(P)}
\end{equation*}
for every family $P$ over $A$ in context $\Gamma$ and every term $x:A$.
\item the judgmental equality
\begin{equation*}
\mftyp{f}\Gamma_0(\subst{x}{Q})\jdeq\subst{\terms[f]{A}(x)}{\mftyp{f}{\ctxext{{\Gamma}{A}}{P}}_0(Q)}
\end{equation*}
for every family $Q$ over $P$ over $A$ in context $\Gamma$ and every term $x:A$.
\end{enumerate}
\end{comment}
\end{enumerate}
\begin{flushright}
\textsl{End of \autoref{defn:premodel-morphism}.}
\end{flushright}
\end{defn}

@@

\begingroup
\color{blue}
\subsubsection*{More desiderata}
Some of which hopefully follow from more elegant or general rules, but
this list is to keep them in mind:
\begin{enumerate}
\item $\id{e_\unit^{-1}(\Gamma.A)}{\msm{e_\unit^{-1}(\Gamma)}{\trans{(\eta_\Gamma)}{A}}}$, where
$\eta_\Gamma:\id{\Gamma}{e_\unit(e_\unit^{-1}(\Gamma))}$ is the unit of the
equivalence $e_\unit$.  
\item $\id{(\msm{A}{P}\to B)}{(P\to \ctxwk{A}{B})}$, expressing that $\Sigma$ is
left adjoint to weakening.
\item $\id{(\ctxwk{A}{B}\to P)}{(B\to\mprd{A}{P})}$ expressing that $\Pi$ is
right adjoint to weakening.
\item a term $\Gamma\vdash\idfunc[A]:A\to A$ for every type $A$ in context $\Gamma$.
\item a context $\UU$ for the universe.
\item an identification $\id{\pi^A\circ \pi^P\circ\iota^x}{\pi^{\subst{x}{P}}}$ for every
family $P$ of types over $A$ in context $\Gamma$ and every term $x:A$. In other
words, the diagram
\begin{equation*}
\begin{tikzcd}
\ctxext{{\Gamma}{A}}{P} \ar{r}{\pi^P} & \ctxext{\Gamma}{A} \ar{d}{\pi^A}\\
\ctxext{\Gamma}{\subst{x}{P}} \ar{u}{\iota^x} \ar{r}[swap]{\pi^{\subst{x}{P}}} & \Gamma
\end{tikzcd}
\end{equation*}
should commute.
\begin{comment}
\item For $\Gamma:\ctx$, $\Gamma\vdash A:\type$, $\ctxext{\Gamma}{A}\vdash P
:\type$ and $\ctxext{{\Gamma}{A}}{P}\vdash Q:\type$, $x:A$ and $u:P$
we desire that $\typefont{nat{-}ev{-}subst}(Q,u,x):\id{Q[\iota^x][\tfev(u,x)]}{Q[u][x]}$.
\item A function
\begin{equation*}
\tfcomp:\terms{\mprd{P}{Q}}\to\prd{x:\terms{A}}\terms{\mprd{\subst{x}{P}}{Q[\iota^x]}}
\end{equation*}
which could possibly replace $\tfev$ because it could be that $\tfev$ is
definable in terms of $\tfcomp$. It should satisfy
\begin{equation*}
\id{\trans{\typefont{nat{-}ev{-}subst}(Q,u,x)}{\tfev(\tfcomp(f,x),\tfev(u,x))}}{\tfev(\lambda(\tfev(f,u)),x)},
\end{equation*}
which is a naturality condition. There should be an equivalence between
$\terms{\mprd{A}{P}}$ and a type consisting of those terms of $\prd{x:\terms{A}}\terms{\subst{x}{P}}$ which satisfy
a certain naturality condition. One could hope that by formulating this dependently
enough, it is possible to write down such a naturality in one line.
\end{comment}
\end{enumerate}
\endgroup

\begin{comment}
\begingroup
\color{darkgreen}
\subsubsection*{The problem with evaluation.}
...or why I'm doing such strange things. It seems that evaluation is of a
different nature. At first sight, one might expect that evaluation compares
to $\lambda$-abstraction the same way that first and second projection compare
to the pairing function $\pairr{\blank,\blank}$. However, this is not the case
because evaluation cannot be seen as an inverse to $\lambda$-abstraction. The
$\lambda$-abstraction is of type $\terms{P}\to\terms{\mprd{A}{P}}$ whereas
evaluation is of type $\terms{\mprd{A}{P}}\to\prd{x:\terms{A}}\terms{\subst{x}{P}}$. There
are a few things to note about $\tfev$:
\begin{enumerate}
\item There's no function back so in particular $\tfev$ is not an equivalence.
My guess is that there should be a naturality condition playing a role.
\item Evaluation tells how functions interact with substitution of terms.
\end{enumerate}
So it seems that evaluation is like a variable binder in the same way
that quantifiers are binding variables. If I remember right, Voevodsky mentioned
something like this at the Formal Topology conference in Ljubljana, though I
forgot what exactly he said.
\endgroup
\end{comment}

\begin{comment}
\begingroup
\color{gray}
\subsubsection*{Old stuff below:}
\begin{enumerate}
\item a context morphism $\catid{\Gamma}:\ctxhom{\Gamma}{\Gamma}$ for every
context $\Gamma$.
\item a context morphism $f\circ g:\ctxhom{E}{\Gamma}$ for every two context
morphisms $f:\ctxhom{\Delta}{\Gamma}$ and $g:\ctxhom{E}{\Delta}$. 
\item paths $\id{f\circ\catid{\Delta}}{f}$ and $\id{\catid{\Gamma}\circ f}{f}$
for every $f:\ctxhom{\Delta}{\Gamma}$.
\item a path $\id{f\circ (g\circ h)}{(f\circ g)\circ h}$ for every
$f:\ctxhom{\Delta}{\Gamma}$, $g:\ctxhom{E}{\Delta}$ and $h:\ctxhom{Z}{E}$. 
\item a context morphism $\Gamma_!:\ctxhom{\Gamma}{\unit^\mfM}$ 
for every context $\Gamma$.
\item a path $\id{(\unit^\mfM)_!}{\catid{\unit^\mfM}}$.
\item a path $\id{\Gamma_!\circ f}{\Delta_!}$ for every $f:\ctxhom{\Delta}{\Gamma}$. 
\item a path $\id{A[\catid{\Gamma}]}{A}$ for every object $A$ in context $\Gamma$.
\item a path $\id{A[f][g]}{A[f\circ g]}$ for every object $A$ in context $\Gamma$
and every $f:\ctxhom{\Delta}{\Gamma}$ and $g:\ctxhom{E}{\Delta}$. 
\end{enumerate}
\endgroup
\end{comment}

@@

\begin{comment}
\section{Continuity of functions}
We wish to find a condition under which we have
\begin{equation*}
\eqv{\terms{\mprd{A}{P}}}{\sm{f:\prd{x:\terms{A}}\terms{\subst{x}{P}}}\text{``$f$ is continuous''}}.
\end{equation*}
for all families $P$ over $A$. 
Needless to say, we first need to formulate what we mean by ``$f$ is continuous''.
Thus, we need to find the family $\pairr{p,u,v}\mapsto \id{\trans{p}{u}}{v}$,
which also requires us to define transportation.

Let $P$ be a family over $A$ in context $\Gamma$. Then we can weaken
$\ctxwk{A}{P}$, which is a family over $\ctxwk{A}{A}$ in context $\ctxext{\Gamma}{A}$.
We also have the family $\ctxwk{\ctxwk{A}{A}}{P}$ over $\ctxwk{A}{A}$ in context
$\ctx{\Gamma}{A}$ and thus we can form the family
\begin{equation*}
\ctxwk{\ctxwk{A}{A}}{P}\to\ctxwk{A}{P}.
\end{equation*}
This family plays the role of the family $x,y:A\vdash P(x)\to P(y):\type$.
{\color{red}Explain this by performing the relevant substitution.}
Transportation is a term of the family
\begin{equation*}
\idtypevar{A}\to\ctxwk{\ctxwk{A}{A}}{P}\to\ctxwk{A}{P}
\end{equation*}
obtained by identity elimination, as usual. Thus we obtain a term
\begin{equation*}
\mathcal{T}rans_{x,y}:\terms{\id{x}{y}\to\subst{x}{P}\to\subst{y}{P}}.
\end{equation*}
for every $x,y:A$.

Now let $x,y:A$. Then we have the type $\subst{y}{P}$
in context $\Gamma$ and the family $\ctxwk{\subst{y}{P}}{\subst{x}{P}}$
over $\subst{y}{P}$ in context $\Gamma$. We could weaken $\subst{y}{P}$ subsequently by $A$, $\ctxwk{A}{A}$
and $\idtypevar{A}$ to obtain the family $\ctxwk{\idtypevar{A}}{\ctxwk{\ctxwk{A}{A}}{\ctxwk{A}{\subst{y}{P}}}}$ over $\idtypevar{A}$ in context $\ctxext{{\Gamma}{A}}{\ctxwk{A}{A}}$. We will
denote this family simply by $\ctxwk{\idtypevar{A}}{\subst{y}{P}}$. Likewise,
we obtain the family $\ctxwk{\idtypevar{A}}{\ctxwk{\subst{y}{P}}{\subst{x}{P}}}$ over
$\ctxwk{\idtypevar{A}}{\subst{y}{P}}$ in context $\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}$ and we obtain the family $\ctxwk{\idtypevar{A}}{\idtypevar{\subst{y}{P}}}$
over $\ctxwk{\idtypevar{A}}{\ctxwk{\subst{y}{P}}{\subst{y}{P}}}$. Now we
have the family $\subst{\ctxwk{A}{\mathcal{T}rans_{x,y}}}{\ctxwk{\idtypevar{A}}{\idtypevar{\subst{y}{P}}}}$ over
$\ctxwk{\idtypevar{A}}{\ctxwk{\subst{y}{P}}{\subst{x}{P}}}$, which is
the desired family to state the continuity condition:

\begin{defn}
A function $f:\prd{x:\terms{A}}\terms{\subst{x}{P}}$ is said to be \emph{continuous}
if there is a term of type
\begin{equation*}
\prd{x,y:\terms{A}}\terms{\subst{\ctxwk{A}{\mathcal{T}rans_{x,y}}}{\ctxwk{\idtypevar{A}}{\idtypevar{\subst{y}{P}}}}}.
\end{equation*}
\end{defn}

\begin{lem}
If all functions are ``continuous mappings of terms'' then types without terms
are empty, i.e.\ we have
\begin{equation*}
(\terms{A}\to\emptyt)\to\terms{A\to\emptyt^\mfGraph}
\end{equation*}
\end{lem}

\begin{proof}
Any function from a type without terms is vacuously continuous.
\end{proof}

We will see that in the graph model functions are not just continuous mappings
of terms, since there is a graph $\tilde{\emptyt}$ which has no terms but which
differs from the empty graph nevertheless.

\begin{desiderata}
The following may relate to functions being continuous mappings of terms (it
seems to be a very strong condition):
\begin{enumerate}
\item We have the adjunctions $\tfcolim\dashv\Delta\dashv\tflim$ in which the
two monads are idempotent (we don't even have the adjunctions for $\mfGraph$).
\item $\eqv{\terms{\Omega}}{\ctx}$
\end{enumerate}
We also need to check whether all functions in the reflexive graphs have this
property (if not, ditch this section).
\end{desiderata}
\end{comment}

>>>>

\subsection{Morphisms of internal models}
\begin{defn}\label{defn:premodel-morphism}
A morphism $f:\mfM\to \mfN$ of internal models consists of
\begin{enumerate}
\item a function $\ctx(f):\ctx(\mfM)\to\ctx(\mfN)$. The function $\ctx(f)$ is also
denoted by $f_0$.
\item a morphism $\mftyp{f}{\Gamma}:\mftyp{\mfM}{\Gamma}\to\mftyp{\mfN}{f_0(\Gamma)}$ of internal models for every
$\Gamma:\ctx(M)$.
\item the judgmental equality
\begin{equation*}
\mftyp{\mftyp{f}{\Gamma}}{A}\jdeq\mftyp{f}{\ctxext{\Gamma}{A}}
\end{equation*}
\item a function $\terms[f]{A}:\terms[\mfM]{A}\to\terms[\mfN]{\mftyp{f}{\Gamma}_0(A)}$ for
every type $A$ in context $\Gamma$. 
\item preservation of context extension: 
\begin{align*}
\alpha^f_0 & :\id{f\circ\tfext^\Gamma}{\tfext^{f_0(\Gamma)}\circ\mftyp{f}{\Gamma}}\\
\alpha^f_1 & :\id{\mftyp{f}{\Gamma}_0\circ\tfext^A}{\tfext^{\mftyp{f}{\Gamma}_0(A)}\circ\mftyp{f}{\ctxext{\Gamma}{A}}}.
\end{align*}
\item preservation of weakening: 
\begin{align*}
\beta^f_0 & :\id{\mftyp{f}{\Gamma}\circ\tfwk^\Gamma}{\tfwk^{f_0(\Gamma)}\circ f}\\
\beta^f_1 & :\id{\mftyp{f}{\ctxext{\Gamma}{A}}\circ\tfwk^A}{\tfwk^\protect{\mftyp{f}{\Gamma}_0(A)}\circ\mftyp{f}{\Gamma}}.
\end{align*}
\item preservation of substitution: 
\begin{equation*}
\gamma^f:\id{\mftyp{f}\Gamma_0\circ\tfsubst^x}{\tfsubst^\protect{\terms[f]{A}(x)}\circ\mftyp{f}{\ctxext{\Gamma}{A}}}.
\end{equation*}
\end{enumerate}
\begin{flushright}
\textsl{End of \autoref{defn:premodel-morphism}.}
\end{flushright}
\end{defn}

@@

\begingroup
\color{blue}
\subsubsection*{More desiderata}
Some of which hopefully follow from more elegant or general rules, but
this list is to keep them in mind:
\begin{enumerate}
\item $\id{e_\unit^{-1}(\Gamma.A)}{\msm{e_\unit^{-1}(\Gamma)}{\trans{(\eta_\Gamma)}{A}}}$, where
$\eta_\Gamma:\id{\Gamma}{e_\unit(e_\unit^{-1}(\Gamma))}$ is the unit of the
equivalence $e_\unit$.  
\item $\id{(\msm{A}{P}\to B)}{(P\to \ctxwk{A}{B})}$, expressing that $\Sigma$ is
left adjoint to weakening.
\item $\id{(\ctxwk{A}{B}\to P)}{(B\to\mprd{A}{P})}$ expressing that $\Pi$ is
right adjoint to weakening.
\item a term $\Gamma\vdash\idfunc[A]:A\to A$ for every type $A$ in context $\Gamma$.
\item a context $\UU$ for the universe.
\item an identification $\id{\pi^A\circ \pi^P\circ\iota^x}{\pi^{\subst{x}{P}}}$ for every
family $P$ of types over $A$ in context $\Gamma$ and every term $x:A$. In other
words, the diagram
\begin{equation*}
\begin{tikzcd}
\ctxext{{\Gamma}{A}}{P} \ar{r}{\pi^P} & \ctxext{\Gamma}{A} \ar{d}{\pi^A}\\
\ctxext{\Gamma}{\subst{x}{P}} \ar{u}{\iota^x} \ar{r}[swap]{\pi^{\subst{x}{P}}} & \Gamma
\end{tikzcd}
\end{equation*}
should commute.
\end{enumerate}
\endgroup

@@
